\toolname: file-based localization of numerical perturbations in data analysis pipelines

### 2020-04-13 paper.tex
Edits in figures

\toolname identifies the components in a pipeline, at the resolution level of a system process, that produce different
results in different execution conditions. First, a directed
bipartite provenance graph is recorded for each pipeline execution, where nodes
represent application processes and files, and edges represent read and
write file accesses (Figure~\ref{fig:provenance-graph}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified and
disambiguated, resulting in a provenance DAG (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Figure~\ref{fig:provenance-dag}). 
DAGs produced in different conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Figure~\ref{fig:labeled-dag}).
 The resulting labeled graph identifies the
non-reproducible processes in the pipeline. \note{Greg}{clarify why many processes don't appear in the graph as shown} \tristan{Ali, 
do you pre-process the reprozip trace to remove some processes, or does the figure show all the processes in this trace?}

@@

We randomly selected 20 unprocessed subjects 
from the HCP data release S500 
available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. \tristan{Ali, I can't find any S500 
data release there, please confirm}
For each subject, available data consisted of 1 or 2 T1w images and 1 or 2 T2w images, with 
$256 \times 320 \times 320$ voxels of size $0.7 \times 0.7 \times 0.7$ mm.
Acquisition protocols and parameters are detailed in~\cite{van2013wu}.

@@

We did not observe any within-OS difference in PreFreeSurfer. In
FreeSurfer, we identified 2 processes leading to within-OS differences due
to the use of pseudo-random numbers: image registration with
\texttt{mri\_segreg}, and cortical surface curvature estimations with
\texttt{mris\_curvature}. Fixing the random seed used in FreeSurfer removed
these differences. \note{Greg}{can you make some statement on the magnitude of these differences?} \tristan{Ali, that's for you.}

@@

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
The processes identified as non-reproducible were observed in linear registration 
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL \texttt{maths}  (in Anatomical Average). 
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing 
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\begin{minipage}{\linewidth}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps (N=20). 
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  number of subjects where the corresponding process appears. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects. \note{Greg}{colorbar doesn't match the numbers: it's from 0-1 but the numbers are 
  subject counts. I don't really have a preference on which, but harmonizing them to be the same unit should happen.} \tristan{
    Ali, it looks like this figure is still unclear, and that other people may still misunderstand it. To better 
    show that the number is a number of subjects, I'd suggest to replace number "x" with "(n=x)",
    write it in smaller font size, and at the bottom-right of the cell instead of in the middle. 
  }
        }
  \label{fig:pfs_heatmap}
\end{minipage}

@@

\begin{minipage}{\linewidth}
  \centering
    \includegraphics[width=\linewidth]{figures/pfs-labeled.pdf} 
    \caption{A complete provenance graph from the PreFreesurfer pipeline.}
    \label{fig:complete_pfs}
\end{minipage}
\end{figure*}

To illustrate the magnitude of the differences,
Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject. Differences appear to be very substantial, in
particular in the frontal and parietal lobes.\note{Greg}{This is very hard to see from the figure because the
luminosity of the two colormaps are so different - the contrast moving from orange to grayscale is difficult. Perhaps use two colors, green and purple for instance, with the same intensity to see if this is more easily seen?} 
\tristan{Ali, I can't find the script to generate this figure in the repo. Could you fix the colors and contrast as per Greg's suggestion?}
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png} 
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs 
    CentOS7). An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here}.}
    \label{fig:fnirt_result}
\end{figure}

>>>>

\toolname identifies the components in a pipeline, at the resolution level of a system process, that produce different
results in different execution conditions. First, a directed
bipartite provenance graph is recorded for each pipeline execution, where nodes
represent application processes and files, and edges represent read and
write file accesses (Figure~\ref{fig:provenance-graph}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified and
disambiguated, resulting in a provenance DAG (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Figure~\ref{fig:provenance-dag}). 
DAGs produced in different conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Figure~\ref{fig:labeled-dag}).
 The resulting labeled graph identifies the
non-reproducible processes in the pipeline. \note{Greg}{clarify why many processes don't appear in the graph as shown} \tristan{Ali, 
do you pre-process the reprozip trace to remove some processes, or does the figure show all the processes in this trace?}
\note{Ali}{It only shows processes that belong to the pipeline (e.g. /usr/local/src/freesurfer/) 
by making a query on the traced DB file. 
We exclude processes like [cp, awk, sed, tee, date\dots]. \tristan{Several points here: (1) freesurfer is obviously not used in this example, so what filtering was actually done? (2) where is this implemented in the code? (3) 
how do you know that you're not breaking the graph by removing these processes?}}

@@

We randomly selected 20 unprocessed subjects 
from the HCP data release S500 
available in \href{https://db.humanconnectome.org}{the ConnectomDB repository} as a subset of the 1200 Subject Release.
For each subject, available data consisted of 1 or 2 T1w images and 1 or 2 T2w images, with 
$256 \times 320 \times 320$ voxels of size $0.7 \times 0.7 \times 0.7$ mm.
Acquisition protocols and parameters are detailed in~\cite{van2013wu}.

@@

We did not observe any within-OS difference in PreFreeSurfer. In
FreeSurfer, we identified 2 processes leading to within-OS differences due
to the use of pseudo-random numbers: image registration with
\texttt{mri\_segreg}, and cortical surface curvature estimations with
\texttt{mris\_curvature}. Fixing the random seed used in FreeSurfer removed
these differences. 
% \note{Greg}{can you make some statement on the magnitude of these differences?} \tristan{Ali, that's for you.}
% \note{Ali}{
% \texttt{mris\_curvature} produces surfaces like lh.inflated.K and lh.inflated.H, 
% I only can see differences using \texttt{mri\_diff} command that it shows:

%   Volumes differ in pixel data

%   maxdiff   0.18604827 at 130715 0 0 0

% for lh.inflated.H and

%   Volumes differ in pixel data

%   maxdiff   0.85469627 at 130682 0 0 0

% for lh.inflated.K

% If you know other ways to check magnitude of differences please let me know.
% \texttt{mri\_segreg} produces different sum files (ASCII file in which summary statistics are saved), 
% but differences belong to meta-data like hostname, date, and SynthSeed. We should remove \texttt{mri\_segreg}
% because it doesn't create data differences.
% The outputs of both executables are in Consider:/data/asalari/ali-tests/paper\_images/
% }

@@

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
The processes identified as non-reproducible were observed in linear registration 
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL \texttt{maths}  (in Anatomical Average). 
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing 
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\begin{minipage}{\linewidth}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps.
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  fraction of subjects for which the corresponding process wasn't reproducible. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects. }
  \label{fig:pfs_heatmap}
\end{minipage}

@@

\begin{minipage}{\linewidth}
  \centering
    \includegraphics[width=\linewidth]{figures/pfs-labeled.pdf} 
    \caption{A complete provenance graph from the PreFreesurfer pipeline. \tristan{Ali, the process tree needs to be in dashed and the file
    dependencies in plain.}}
    \label{fig:complete_pfs}
\end{minipage}
\end{figure*}

To illustrate the magnitude of the differences,
Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject. Differences appear to be very substantial, in
particular in areas framed in red.
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png} 
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs 
    CentOS7). An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here}.}
    \label{fig:fnirt_result}
\end{figure}
